---
title: 'HW 5: Correlations and linear models. Tests for categorial variables'
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(vcd) # for mosaic plots
```

## 1. Vowel reduction in Russian
Pavel Duryagin ran an experiment on perception of vowel reduction in Russian language. The dataset `shva` includes the following variables:  
_time1_ - reaction time 1  
_duration_ - duration of the vowel in the stimuly (in milliseconds, ms)  
_time2_ - reaction time 2  
_f1_, _f2_, _f3_ - the 1st, 2nd and 3rd formant of the vowel measured in Hz (for a short introduction into formants, see [here](https://home.cc.umanitoba.ca/~krussll/phonetics/acoustic/formants.html))  
_vowel_ - vowel classified according the 3-fold classification (_A_ - _a_ under stress, _a_ - _a/o_ as in the first syllable before the stressed one, _y_ (stands for shva) - _a/o_ as in the second etc. syllable before the stressed one or after the stressed syllable, cf. _g_[_y_]_g_[_a_]_t_[_A_]_l_[_y_] _gogotala_ `guffawed').  
In this part, we will ask you to analyse correlation between f1, f2, and duration.
The dataset is available [https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/duryagin_ReductionRussian.txt](here).

### 1.0 
Read the data from file to the variable `shva`.
```{r 1.0}
shva <- read_delim("https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/duryagin_ReductionRussian.txt", delim="\t")
```

```{r}
shva
```

### 1.1 
Scatterplot `f1` and `f2` using `ggplot()`. 
Design it to look like the [following](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin1.png).
```{r 1.1}
ggplot(shva, aes(x = f2, y = f1, color = factor(vowel))) +
  geom_point()
```

### 1.2 
Plot the boxplots of `f1` and `f2` for each vowel using `ggplot()`.
Design it to look like [this](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin2.png) and [this](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin3.png).
```{r 1.2}
# f1 boxplot
ggplot(data = shva, aes(x = f1, y = vowel, fill = vowel)) +
  geom_boxplot()
# f2 boxplot
ggplot(data = shva, aes(x = f2, y = vowel, fill = vowel)) +
  geom_boxplot()
```

### 1.3 
Calculate Pearson's correlation of `f1` and `f2` (all data)
```{r 1.3}
cor.test(shva$f1, shva$f2, method=c("pearson"))
```


### 
1.4 Calculate Pearson's correlation of `f1` and `f2` for each vowel
```{r 1.4}
cor.test(filter(shva, shva$vowel == "A")$f1, filter(shva, shva$vowel == "A")$f2, method=c("pearson"))
cor.test(filter(shva, shva$vowel == "a")$f1, filter(shva, shva$vowel == "a")$f2, method=c("pearson"))
cor.test(filter(shva, shva$vowel == "y")$f1, filter(shva, shva$vowel == "y")$f2, method=c("pearson"))
```

## 2 Linear regressions

### 2.1.1 
Use the linear regression model to predict `f2` by `f1`.
```{r 2.1.1}
linearMod <- lm(data=shva, f2 ~ f1) 
summary(linearMod)
```

### 2.1.2 
Write down the equation for f2 using coefficients from the model (e.g. $y =  b + kx$)
```{r 2.1.2}
shva$f2 = 1639.7021 -0.4287 * shva$f1
```

### 2.1.3 
Provide the adjusted R$^2$
```{r 2.1.3}
summary(linearMod)$adj.r.squared 
```

### 2.1.4 
Add the regression line in the scatterplot 1.1.
```{r 2.1.4}
scatter.smooth(x=shva$f1, y=shva$f2, main="f2 ~ f1") 
```

### 2.1.5 
Make a scatter plot for `f1` and `f2` grouped by vowels. 
Use `ggplot()` and `facet_wrap()`.

```{r 2.1.5}

```

### 2.2.1 
Use the linear regression model to predict `f2` by `f1` and `vowel`.
```{r 2.2.1}
trainingRowIndex <- sample(1:nrow(shva), 0.8*nrow(shva))  # row indices for training data
trainingData <- shva[trainingRowIndex, ]  # model training data
testData  <- shva[-trainingRowIndex, ]   # test data
linearMod <- lm(f2 ~ f1 + vowel, data=trainingData)
Predicts <- predict(linearMod, testData)
summary(linearMod)
actuals_preds <- data.frame(cbind(actuals=testData$dist, predicteds=Predicts))  # make actuals_predicteds dataframe.
correlation_accuracy <- cor(actuals_preds)  
print(correlation_accuracy)
```

### 2.2.2 
What is the intercept of the model?
```{r 2.2.2}
summary(linearMod)$coefficients["(Intercept)", "Estimate"]
```

### 2.2.3 
Provide the adjusted R$^2$
```{r 2.2.3}
summary(linearMod)$adj.r.squared 
```

### 2.2.4 
Write down your general conclusions about the relationship between `f1`, `f2`, and `vowels`.

```

```

## 3. Dutch causative constructions

When the Dutch use two near-synonymous periphrastic causative verbs, *doen* and *laten*?
```
       De politie deed/liet de auto stoppen.
  lit. the police did/let the car stop 
       'The police stopped the car'
```

This is a data set on two rival constructions with *doen* and *laten* sampled from the newspaper corpora. The data frame includes 500 observations on the following 7 variables:

* `Aux` -- verb: a factor with levels `doen` and `laten`  
* `CrSem` -- the semantic class of the Causer: a factor with levels `Anim` (animate) and `Inanim` (inanimate)  
* `CeSem` -- the semantic class of the Causee: a factor with levels `Anim` (animate) and `Inanim` (inanimate)  
* `CdEvSem` -- the semantic domain of the caused event expressed by the Effected Predicate: a factor with levels `Ment` (mental) and `NonMent` (e.g. physical or social)  
* `CeSynt` -- the syntactic status of the Causee: a factor with levels `Clause`, `Impl` (implicit, not expressed), `NP` (noun phrase), `PP` (prepositional phrase)    
* `EPTrans` -- transitivity or intransitivity of the effected predicate, a factor with two levels `Tr` and `Intr`  
* `Country` -- a factor with levels `BE` (Belgium) and `NL` (Netherlands)  
* `Domain` -- a factor with four levels for newspaper domains.    

Data from Natalya Levshina's `RLing` package available (here)[https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/dutch_causatives.csv]
Read more on the constructions in [Levhina, Geerarts, Speelman 2014](https://www.academia.edu/7287585/Dutch_causative_constructions_Quantification_of_meaning_and_meaning_of_quantification_with_Dirk_Geeraerts_and_Dirk_Speelman_).

### 3.0 
Read the data from file to the variable `d_caus`.
```{r 3.0}
d_caus <- read.csv("https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/dutch_causatives.csv")
summary(d_caus)
```
```{r}
d_caus
```

### 3.1 
We are going to test whether the association between `Aux` and other categorical variables (`Aux` ~ `CrSem`, `Aux` ~ `CeSem`, etc) is statistically significant. The assiciation with which variable should be analysed using Fisher's Exact Test and not using Pearson's Chi-squared Test? Is this association statistically significant?

Answer: Fisher's Exact Test is used for contigency table 2X2 (when we examine 2 variables with 2 levels each), Pearson's Chi-squared Test - for any contigency tables
```{r 3.1}
chisq.test(d_caus$CeSynt, d_caus$Aux)
chisq.test(d_caus$Domain, d_caus$Aux)
fisher.test(d_caus$CrSem, d_caus$Aux)
fisher.test(d_caus$CeSem, d_caus$Aux)
fisher.test(d_caus$CdEvSem, d_caus$Aux)
fisher.test(d_caus$EPTrans, d_caus$Aux)
fisher.test(d_caus$Country, d_caus$Aux)
```

### 3.2. 
Test the hypothesis that `Aux` and `EPTrans` are not independent with the help of Pearson's Chi-squared Test. 
```{r 3.2}
EPT_chi <- chisq.test(d_caus$EPTrans, d_caus$Aux)
```

### 3.3 
Provide expected frequencies for Pearson's Chi-squared Test of `Aux` and `EPTrans` variables.
```{r 3.3}
EPT_chi$expected
```

### 3.4. 
Calculate the odds ratio for observed frequencies of `Aux` and `EPTrans`
For 2×2 contigency table
$$\begin{matrix}
a & b \\
c & d
\end{matrix}
$$
one can find *odds ratio* as $(a/c)/(b/d)$.

```{r 3.4}
EPT_chi$observed
(EPT_chi$observed[1,1] / EPT_chi$observed[2,1]) / (EPT_chi$observed[1,2] / EPT_chi$observed[2,2])

```

### 3.4.1 
Find odds ratio for expected frequencies of `Aux` and `EPTrans`

```{r 3.4.1}
(EPT_chi$observed[1,1] / EPT_chi$observed[2,1]) / (EPT_chi$observed[1,2] / EPT_chi$observed[2,2])
```
What can you say about odds ratio of expected frequencies for arbitrary data?

### 3.5 
Calculate effect size for this test using Cramer's V (phi).
```{r 3.5}
library(lsr)
cramersV(d_caus$EPTrans, d_caus$Aux)
```

### 3.6. 
Report the results of independence test using the following template:
```
We have / have not found a significant association between variables ... and ... (p < 0.001).  The odds of ... were ... times higher / lower in (group ....) than in (group ....). Effect size is large / medium / small (Cramer's V = ....).
```
We have not found a significant association between variables Aux and EPTrans (p < 0.001).  The odds of "doen" were 2.606162 times higher in Intr than in Tr. Effect size is small (Cramer's V = 0.1691587).

### 3.7 
Visualize the distribution using mosaic plot.
Use `mosaic()` function from `vcd` library.
```{r 3.7}
mosaic(~ Aux + EPTrans, data=d_caus, shade=TRUE, legend=TRUE)
```

Below is an example of how to use mosaic() with three variables.
```{r 3.7.1}
# mosaic(~ Aux + CrSem + Country, data=d_caus, shade=TRUE, legend=TRUE)
```

### 3.8 
Why is it not recommended to run multiple Chisq tests of independence on different variables within your dataset whithout adjusting for the multiplicity? (i.e. just testing all the pairs of variables one by one)  
```
It can be a bad idea because every time we run Chisq test the chance that it return an error increase. So the more tests we make the bigger the probabolity of an error becomes so it is more and more possible that the test is incorrect

```

### 3.9 
Provide a short text (300 words) describing the hypothesis of this study and the results of your analysis.
```
One can find two semantically very close causative verbs in the Dutch language they are "doen" and "laten". The hypothesys tested in this research is that there are any linguistic reasons of their existance. The dataset for the research contains various linguistic information about the context of these two words (semantic class of the Causer, the semantic class of the Causee and so on...)
It was decided to check the hypothesis using, firstly, the Fisher's exact test to test the associations Aux ~ CeSynt and Aux ~ EPTrans. By it we found out that there is an affection of Causee on the choice of the verbs. Later we applied the Pearson's Chi-squared Test and discovered the verb "doen" is more frequently associated with intransitive effected verbs comparing to "laten" what we cannot say about Aux and EPTrans
To sum up the hypothesis was proved and despite the fact  "doen" and "laten" may look almost full synonyms they are met in different contexts

```
